Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2018-09-03 Response_to_Reviewers.tex

Concerning specifically the comments made by the first referee:

\begin{itemize}
	% 1. I think it would be better if the author would focus on a single model. This way they could describe the model and the analyses in more detail. As it stands, the models are treated as black boxes. 
	\item This is indeed a qualified comment. However, after presenting our research to several audiences, we came to the conclusion that the genericity of our approach would be better served by two case studies than with a single model. We acknowledge the fact that this choice led us to shorten their presentations. Nevertheless, we have been taking heed to your remark to improve the description of the models by providing hyperlinks towards the NetLogo pages and Scala code lines (to be done). We believe this is sufficient as the models have not been adjusted from these textbook cases implementations.

   % I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
   \item

	% 2. It is very important to describe the models. This is especially relevant for the Schelling model, which has many variants; each has different dependencies on the initial pattern. In my opinion, the NetLogo implementation is not the best choice because the model does not reach a state of segregation when agents are highly intolerant.
	\item \comment[MLT]{Thank you for this comment. The reason why we have decided to use the NetLogo implementations was again to help the readers to distance themselves from the case studies and focus on the genericity of the approach. This also explains why we have chosen Sugarscape as a second example: these two models are often used as toy models due to the simplicity of their mechanisms (in the NetLogo scenarios) and therefore do not require a high level of expertise nor technicality to apprehend the impacts of the initial spatial conditions on the model outputs. This is again an argument in favour of our choice: space-dependent variations are to be expected even with the most simple sociospatial models.}

### 2018-09-03 paper.tex
compiles
.

%\corrauth{Juste Raimbault, UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
%13 rue du Four,
%75006 Paris, France.
%}
%\corrauth{}

@@

\begin{abstract}
%\noindent
%\setlength{\parindent}{0pt}
Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions. Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment. In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create density grids with which spatial simulation models are initialised. We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant in the two models. Furthermore, this effect is sometimes larger than the effect of parameters' value change. 
\end{abstract}

@@

%\maketitle

%% main text

\enlargethispage{-15mm}

@@

%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams}
\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).}
\label{fig:sugarscape-distance-meta}
\end{figure}
%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/relativedistance_morphspace}
\caption{\textbf{Relative distances of phase diagrams to the reference across grids.} Relative distance as a function of two first principal components of the morphological space (see text). Black point correspond to the reference spatial configuration. Green frame and blue frame give respectively the first and second particular phase diagrams shown in Fig.~\ref{fig:sugarscape-phasediagrams}.}
\label{fig:sugarscape-distance-pca}
\end{figure}
%%%%%%%%%%%%%

@@

\begin{table*}[]
\centering
\begin{threeparttable}
\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.}}
\label{tab:regressionSchelling}
%\begin{adjustwidth}{-1cm}{-1cm}
\begin{tabular}{|m{2.5cm}|ll|ll|ll|}
\hline
Simulation outcome by segregation index:    & \multicolumn{2}{c|}{\textbf{Dissimilarity}}   & \multicolumn{2}{c|}{\textbf{Entropy}} & \multicolumn{2}{c|}{\textbf{Moran's I}} \\ \hline
\textbf{Intercept}                          & -0.212 *** & -0.141 ***                       & -0.254 ***        & -0.208 ***        & -0.036 ***           & -0.061 ***               \\ \hline
\textbf{Similarity Wanted (S)}              & 1.212 ***  & 1.212 ***                        & 1.250 ***         & 1.250 ***         & 0.550 ***            & 0.550 ***                \\ 
\textbf{quadratic term ($S^2$)}               & -0.942 *** & -0.942 ***                       & -0.963 ***        & -0.963 ***        & -0.428 ***           & -0.438 ***               \\ 
\textbf{Vacancy Rate (V)}                   & 0.602 ***  & 0.602 ***                        & 0.453 ***         & 0.453 ***         & -0.027 ***           & -0.027 ***               \\ 
\textbf{Minority Index (\%Maj - \%Min)}     & 0.307å ***  & 0.307 ***                        & 0.130 ***         & 0.130 ***         & -0.067 ***           & -0.067 ***               \\ \hline
\textbf{Density Grid = Polycentric}         &            & 0.087 ***                        &                   & 0.052 ***         &                      & 0.001 ***                \\ 
\textbf{Density Grid = Discontinuous}       &            & 0.111 ***                        &                   & 0.068 ***         &                      & \textit{0.00}              \\
\textbf{Attraction meta-parameter $\alpha$} &            & -0.083 ***                       &                   & -0.053 ***        &                      & 0.014 ***                \\ 
\textbf{Diffusion meta-parameter $\beta$}   &            & 0.323 ***                        &                   & 0.218 ***         &                      & 0.017 ***           \\ \hline
\textbf{R2 (\%)}                            & 30.6       & 34.7                             & 24.1              & 25.6              & 23.9                 & 24.0                    \\ 
\textbf{\# of observations (sim. runs)}     & 2,106,000  & 2,106,000 						 & 2,106,000          & 2,106,000          & 2,106,000             & 2,106,000                \\ 
\textbf{AIC}                                & -70717.68   & -198748.2  						& 208213.8          & 166048.8          & -4385990             & -4387816                 \\ \hline
\end{tabular}
%\end{adjustwidth}
\begin{tablenotes}
 \item Moran's I applies to the minority population
 \item *** means that the estimate is significant at the 0.01 level.
\end{tablenotes}
  \end{threeparttable}
\end{table*}

@@

%%%%%%%%%%%%%
\begin{figure}
\centering
\includegraphics[width=0.48\textwidth]{figures/phasediagram_id27_maxSugar110}
\includegraphics[width=0.48\textwidth]{figures/phasediagram_id0_maxSugar110}
\caption{\textbf{Examples of phase diagrams.} We show two dimensional phase diagrams on $(P,s_-)$, both at fixed $s_+ = 110$. (Left) Green frame, obtained with $\alpha = 0.79$, $n=2$, $\beta = 0.14$, $N=157$; (Right) Blue frame, obtained with $\alpha = 2.56$, $n=3$, $\beta = 0.13$, $N=128$.}
\label{fig:sugarscape-phasediagrams}
\end{figure}
%%%%%%%%%%%%%

>>>>

@@

\begin{abstract}
Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions. Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment. In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create density grids with which spatial simulation models are initialised. We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant in the two models. Furthermore, this effect is sometimes larger than the effect of parameters' value change. 
\end{abstract}

@@

@@

%%%%%%%%%%%%%
\begin{figure*}
\centering
\includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams}
\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).}
\label{fig:sugarscape-distance-meta}
\end{figure*}
%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%
\begin{figure*}
\centering
\includegraphics[width=0.8\textwidth]{figures/relativedistance_morphspace}
\caption{\textbf{Relative distances of phase diagrams to the reference across grids.} Relative distance as a function of two first principal components of the morphological space (see text). Black point correspond to the reference spatial configuration. Green frame and blue frame give respectively the first and second particular phase diagrams shown in Fig.~\ref{fig:sugarscape-phasediagrams}.}
\label{fig:sugarscape-distance-pca}
\end{figure*}
%%%%%%%%%%%%%

@@

\begin{table*}[]
\centering
\begin{threeparttable}
\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.}}
\label{tab:regressionSchelling}
%\begin{adjustwidth}{-1cm}{-1cm}
\begin{tabular}{|p{2.5cm}|ll|ll|ll|}
\hline
Simulation outcome by segregation index:    & \multicolumn{2}{c|}{\textbf{Dissimilarity}}   & \multicolumn{2}{c|}{\textbf{Entropy}} & \multicolumn{2}{c|}{\textbf{Moran's I}} \\ \hline
\textbf{Intercept}                          & -0.212 *** & -0.141 ***                       & -0.254 ***        & -0.208 ***        & -0.036 ***           & -0.061 ***               \\ \hline
\textbf{Similarity Wanted (S)}              & 1.212 ***  & 1.212 ***                        & 1.250 ***         & 1.250 ***         & 0.550 ***            & 0.550 ***                \\ 
\textbf{quadratic term ($S^2$)}               & -0.942 *** & -0.942 ***                       & -0.963 ***        & -0.963 ***        & -0.428 ***           & -0.438 ***               \\ 
\textbf{Vacancy Rate (V)}                   & 0.602 ***  & 0.602 ***                        & 0.453 ***         & 0.453 ***         & -0.027 ***           & -0.027 ***               \\ 
\textbf{Minority Index (\%Maj - \%Min)}     & 0.307å ***  & 0.307 ***                        & 0.130 ***         & 0.130 ***         & -0.067 ***           & -0.067 ***               \\ \hline
\textbf{Density Grid = Polycentric}         &            & 0.087 ***                        &                   & 0.052 ***         &                      & 0.001 ***                \\ 
\textbf{Density Grid = Discontinuous}       &            & 0.111 ***                        &                   & 0.068 ***         &                      & \textit{0.00}              \\
\textbf{Attraction meta-parameter $\alpha$} &            & -0.083 ***                       &                   & -0.053 ***        &                      & 0.014 ***                \\ 
\textbf{Diffusion meta-parameter $\beta$}   &            & 0.323 ***                        &                   & 0.218 ***         &                      & 0.017 ***           \\ \hline
\textbf{R2 (\%)}                            & 30.6       & 34.7                             & 24.1              & 25.6              & 23.9                 & 24.0                    \\ 
\textbf{\# of observations (sim. runs)}     & 2,106,000  & 2,106,000 						 & 2,106,000          & 2,106,000          & 2,106,000             & 2,106,000                \\ 
\textbf{AIC}                                & -70717.68   & -198748.2  						& 208213.8          & 166048.8          & -4385990             & -4387816                 \\ \hline
\end{tabular}
%\end{adjustwidth}
\begin{tablenotes}
 \item Moran's I applies to the minority population
 \item *** means that the estimate is significant at the 0.01 level.
\end{tablenotes}
  \end{threeparttable}
\end{table*}

@@

%%%%%%%%%%%%%
\begin{figure*}
\centering
\includegraphics[width=0.48\textwidth]{figures/phasediagram_id27_maxSugar110}
\includegraphics[width=0.48\textwidth]{figures/phasediagram_id0_maxSugar110}
\caption{\textbf{Examples of phase diagrams.} We show two dimensional phase diagrams on $(P,s_-)$, both at fixed $s_+ = 110$. (Left) Green frame, obtained with $\alpha = 0.79$, $n=2$, $\beta = 0.14$, $N=157$; (Right) Blue frame, obtained with $\alpha = 2.56$, $n=3$, $\beta = 0.13$, $N=128$.}
\label{fig:sugarscape-phasediagrams}
\end{figure*}
%%%%%%%%%%%%%
